---
title: "Legislative Endurance"
author: "Brent Schneeman"
date: "March 25, 2016"
output: 
  html_document: 
    keep_md: yes
license: CC-BY
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(yaml)
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggalt)
library(ggthemes)

ledge <- yaml.load_file("./data/legislators-historical.yaml")

# want to create a dataframe with id, name, term type, term start, term end, and party out of the ledge object


extractElements <- function(e){
  id <- e$id$govtrack
  firstName <- e$name$first
  lastName <- e$name$last
  gender<-e$bio$gender

  terms = bind_rows(lapply(e$terms, FUN=function(t) data.frame(
    type=t$type, 
    start=t$start, 
    end=t$end, 
    party=if(is.null(t$party)) "Unknown" else t$party, stringsAsFactors = F)))
  
  terms <- terms %>% arrange(start)
  
  # some legislators will jump between the House and the Senate, which is encoded
  # as "rep" or "sen" in terms$type. The RLE function figures out how many terms
  # each "run" of House or Senate lasts.
  
  runs <- rle(terms$type)
  
  # these statements pick out the start and end of each run, along with the parties
  # and the body served (the type).
  
  # Note that it is possible for a legislator to switch parties; I'm not dealing with
  # that. yet.
  
  starts <- terms$start[c(1,cumsum(head(runs$lengths,-1)) +1)]
  ends <- terms$end[cumsum(runs$lengths)]
  parties <- terms$party[cumsum(runs$lengths)]
  types <- terms$type[cumsum(runs$lengths)]
  
  data.frame(id=rep(id, length(starts)),
             firstName = rep(firstName, length(starts)),
             lastName = rep(lastName, length(starts)),
             gender = rep(gender, length(starts)),
             numTerms = runs$lengths,
             start = starts,
             end = ends,
             party = parties,
             type = types,
             stringsAsFactors = FALSE)
}

splinePoints <- function(theIndex, df){
  
  e <- df[theIndex,]
  
  # peak of the spline is the number of terms served
  peak <- e$numTerms

  # this output dataframe defines the shape of the spline. Lots of 
  # customization here.
  data.frame(x=quantile(c(e$start, e$end), probs=c(0,.15,.85,1)), 
             y=c(0, peak*.85, peak*.85, 0),
             splineIndex = e$id,
             party=e$party)
}


temp = lapply(ledge, extractElements)
ledge_df = bind_rows(temp)

ledge_df$start= ymd(ledge_df$start)
ledge_df$end <- ymd(ledge_df$end)
ledge_df$gender <- as.factor(ledge_df$gender)
ledge_df$party <- as.factor(ledge_df$party)
ledge_df$type <- as.factor(ledge_df$type)
ledge_df$dur <- ledge_df$end - ledge_df$start

ledge_df[ledge_df$party=="Republican",]$numTerms <- ledge_df[ledge_df$party=="Republican",]$numTerms * -1

```
```{r Senate, echo=FALSE}

body <- ledge_df %>% 
  filter(dur < 22000) %>%
  filter(type=="sen") %>% 
  filter(party %in% c("Democrat", "Republican"))%>% 
  arrange(start, end) 

# cheap and dirty way to move along a dataframe
temp <- lapply(X=seq_len(nrow(body)), splinePoints, body)
theSplines <- bind_rows(temp)


ggplot(theSplines, aes(x, y, group=splineIndex, colour=party)) + 
  geom_xspline(spline_shape = -0.5, size=0.2) +
  scale_color_manual("US Senate", values=c("blue", "red")) +
  theme_few() + theme(axis.text.y = element_blank(),
                      axis.title.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      panel.border = element_blank()) +
  xlab("Year")
```

## Entrenched Interests

The Founding Fathers created the United States with the concept of citizen servants in mind. Men (originally, and land-owners at that) would represent constituents for a time and then return to their lives. Fresh-off the American Revolution, and having land to tend, this seemed like a good idea: who would want to spend time governing instead of working the land and being a member of the community where you built your life?

After being elected to two terms of the Presidency, George Washington decided to not run again, not wanting to create a democratic monarchy. Soon, the  government codified Presidential Term Limits in the US Constitution. Those term limits never trickled down to the Senate or the House; over time, legislators have spent more and more time legislating at the expense of being a member of their "home" communities.

Today, American Congressional Legislators serve longer terms each decade and have wandered away from the citizen servant ideal. The United States is a much more complex country than it was 230 years ago and perhaps having "professional" representatives is best for the Union. However, the current system of multi-term careers and safe districts leads to entrenched special interests; legislators get elected to careers, not to terms of service to their constituents.

```{r echo=FALSE}
ledge_df$numTerms = abs(ledge_df$numTerms)

body <- ledge_df %>% 
  filter(type=="sen") %>% 
  arrange(start, end) 

# cheap and dirty way to move along a dataframe
temp <- lapply(X=seq_len(nrow(body)), splinePoints, body)
theSplines <- bind_rows(temp)


ggplot(theSplines, aes(x, y, group=splineIndex)) + 
  geom_xspline(spline_shape = -0.5, size=0.2, colour="black") +
  theme_few() + theme(axis.text.y = element_blank(),
                      axis.title.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      panel.border = element_blank()) +
  xlab("Year")

```

This plot shows the service arcs of every Senator to have *finished* serving (that is, currently sitting Senators are not shown). The heights of the arcs are related to the number of terms served the the horizontal axis progresses by year to the right. The arcs tend to get higher over time, which indicates that Senators are serving longer. The horizontal striations - the *levels* - exactly correspond to the number of terms served: one term, two terms, etc, and the levels get higher towards the right edge of the graph.

```{r echo=FALSE}
bks = seq(trunc(min(year(body$end)) / 10) * 10, trunc(max(year(body$end)) / 10) * 10 + 10, 10)
body$decade = bks[cut(year(body$end), breaks=bks, labels=FALSE)]

decades <- body %>%
  group_by(decade) %>% 
  summarize(aveDur = mean(as.numeric(dur, units="days")/365),
            aveTerms = mean(numTerms)) 

ggplot(decades, aes(x=decade, y=aveTerms, colour=decade)) + 
  geom_point(colour="black") + geom_smooth(method="lm", se=F, linetype="dashed", size=0.5) +
  theme_few() + ylab("Average Number of Terms") +
  xlab("Decade in which service ended") +
  theme(panel.border = element_blank()) 

```

A simple scatter plot (with a trend-line) shows the average of the number of terms for every Senator who finished serving in a partcular decade. This clearly shows an up-and-to-the-right increase: the number of terms each Senator is serving is increasing.

## Party time

```{r echo=FALSE, message=FALSE}
belows = c("Republican", "Jackson", "Unknown", "Adams", "American", "Populist")
ldf = ledge_df
ledge_df[ledge_df$party %in% belows,]$numTerms <- ledge_df[ledge_df$party %in% belows,]$numTerms * -1

body <- ledge_df %>% 
  filter(dur < 22000) %>%
  filter(type=="sen") %>% 
  arrange(start, end) 

# cheap and dirty way to move along a dataframe
temp <- lapply(X=seq_len(nrow(body)), splinePoints, body)
theSplines <- bind_rows(temp)

ggplot(theSplines, aes(x, y, group=splineIndex, colour=party)) + 
  geom_xspline(spline_shape = -0.5, size=0.2) +
  theme_few() + theme(axis.text.y = element_blank(),
                      axis.title.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      panel.border = element_blank()) +
  xlab("Year") + labs(colour="Party\nAffiliation")
```

American politics and political parties go hand-in-hand. America has had lots of Senators from across a wide spectrum of parties serve. Trying to break them out visually is difficult, and probably not all that important. While the "Liberal Republican", "Readjuster", and other parties have a place in American History, today the landscape is dominated by two parties: "Democrat" and "Republican". A two-party system has been in-place since just after the Founding. Our "winner-takes-all" election system tends to suppress third-party efforts as coalitions are formed to garner the greatest number of votes.

```{r echo=FALSE}
ledge_df = ldf

ledge_df[ledge_df$party=="Republican",]$numTerms <- ledge_df[ledge_df$party=="Republican",]$numTerms * -1

body <- ledge_df %>% 
  filter(dur < 22000) %>%
  filter(type=="sen") %>% 
  filter(party %in% c("Democrat", "Republican"))%>% 
  arrange(start, end) 

# cheap and dirty way to move along a dataframe
temp <- lapply(X=seq_len(nrow(body)), splinePoints, body)
theSplines <- bind_rows(temp)


ggplot(theSplines, aes(x, y, group=splineIndex, colour=party)) + 
  geom_xspline(spline_shape = -0.5, size=0.2) +
  scale_color_manual("US Senate", values=c("blue", "red")) +
  theme_few() + theme(axis.text.y = element_blank(),
                      axis.title.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      panel.border = element_blank()) +
  xlab("Year")

```

Which brings us to the plot at the top of this post (replicated immediately above). It shows every Democrat and Republican Senator who has served since the inception of the Union (not including those currently serving). The image shows the Republican party has two phases. The Republican label has been with the Union since its inception, and then disappeared around 1830, right about the same time the first Democrat Senators appeared. While a few Republicans came back before 1850 (James Simmons, Hannibal Hamlin, John Hale, and others), the party didn't really show strength until between 1863 and 1870 when 47 Republican Senators were elected: "The Party of Lincoln", indeed. Little connection between the early Republican label and the 1860s party likely exists.

Neither party today, however, is likely similar to their 1860s incarnations. [1933 is the start of the Fifth and Sixth party systems](https://en.wikipedia.org/wiki/Political_parties_in_the_United_States#Fifth_and_Sixth_Party_Systems:_1933-present) and since the 1930s, the average number of terms served is on a steadily increasing slope.

```{r, echo=FALSE}
bks = seq(trunc(min(year(body$end)) / 10) * 10, trunc(max(year(body$end)) / 10) * 10 + 10, 10)
body$decade = bks[cut(year(body$end), breaks=bks, labels=FALSE)]

decades <- body %>% filter(decade>"1930") %>% 
  group_by(decade,party) %>% 
  summarize(aveDur = mean(as.numeric(dur, units="days")/365),
            aveTerms = mean(abs(numTerms))) 

ggplot(decades, aes(x=decade, y=aveTerms, colour=party)) + 
  geom_point() + geom_smooth(method="lm", se=F, linetype="dashed", size=0.5) +
  scale_color_manual("US Senate", values=c("blue", "red")) +
  theme_few() + ylab("Average Number of Terms") +
  theme(panel.border = element_blank()) + coord_cartesian(ylim=c(0,8))
```

The slope of the trendline for the "Democrat" label (0.019) is over twice that of the "Republican" label (0.008), meaning that as each decade passes, a Democratic Senator "gains" twice the longevity as his or her Republican counterpart. For the House of Representatives, the trends look like:

```{r, echo=FALSE}
body <- ledge_df %>% 
  filter(dur < 22000) %>%
  filter(type=="rep") %>% 
  filter(party %in% c("Democrat", "Republican"))%>% 
  arrange(start, end) 

bks = seq(trunc(min(year(body$end)) / 10) * 10, trunc(max(year(body$end)) / 10) * 10 + 10, 10)
body$decade = bks[cut(year(body$end), breaks=bks, labels=FALSE)]

decades <- body %>% filter(decade>"1930") %>% 
  group_by(decade,party) %>% 
  summarize(aveDur = mean(as.numeric(dur, units="days")/365),
            aveTerms = mean(abs(numTerms))) 

ggplot(decades, aes(x=decade, y=aveTerms, colour=party)) + 
  geom_point() + geom_smooth(method="lm", se=F, linetype="dashed", size=0.5) +
  scale_color_manual("US House", values=c("blue", "red")) +
  theme_few() + ylab("Average Number of Terms") +
  theme(panel.border = element_blank()) + coord_cartesian(ylim=c(0,8))

```

Here, the Democrats "enjoy" a more modest trend advantage. Senate elections are statewide and cannot be "gerrymandered" efficienctly. In the House, both parties benefit from "safe" legislative districts that are designed to protect incumbency. The slopes for the House may relfect this as they are steeper - in each decade, Representatives gain more tenure than their Senate counterparts.

## Most Professional

```{r, echo=FALSE, message=FALSE}

body <- ledge_df %>% 
  filter(dur < 22000) %>%
  filter(type=="sen") %>% 
  arrange(start, end) 

body %>% mutate(numTerms = abs(numTerms)) %>% 
  select(firstName, lastName, party, numTerms, start, end, dur) %>%
  arrange(desc(dur)) %>% top_n(10) %>% knitr::kable()

```

The top ten longest serving Senators, presented as FYI.

## Credits, Rights and How-to

The source data for these plots is courtesy of [GovTrack.us](https://www.govtrack.us/), and specifically their [historical legislator](https://github.com/govtrack/congress-legislators) YAML dataset. GovTrack releases any of their copyrighted data into the public domain, under a [CC-BY Creative Commons](https://creativecommons.org/licenses/by/3.0/us/) license; I do the same for the plots and code used to create this.

The code used to generate this document is found in [github.com/schnee/ledge-dur](https://github.com/schnee/ledge-dur), specifically in the "endurance.Rmd" document. You will need the R environment to process it; I recommend [RStudio](https://www.rstudio.com/) if you don't already have a suitable environment.