---
title: "Legislative Endurance"
author: "Brent Schneeman"
date: "March 25, 2016"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(yaml)
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggalt)
library(ggthemes)

ledge <- yaml.load_file("./data/legislators-historical.yaml")

# want to create a dataframe with id, name, term type, term start, term end, and party out of the ledge object


extractElements <- function(e){
  id <- e$id$govtrack
  firstName <- e$name$first
  lastName <- e$name$last
  gender<-e$bio$gender

  terms = bind_rows(lapply(e$terms, FUN=function(t) data.frame(
    type=t$type, 
    start=t$start, 
    end=t$end, 
    party=if(is.null(t$party)) "Unknown" else t$party, stringsAsFactors = F)))
  
  terms <- terms %>% arrange(start)
  
  # some legislators will jump between the House and the Senate, which is encoded
  # as "rep" or "sen" in terms$type. The RLE function figures out how many terms
  # each "run" of House or Senate lasts.
  
  runs <- rle(terms$type)
  
  # these statements pick out the start and end of each run, along with the parties
  # and the body served (the type).
  
  # Note that it is possible for a legislator to switch parties; I'm not dealing with
  # that. yet.
  
  starts <- terms$start[c(1,cumsum(head(runs$lengths,-1)) +1)]
  ends <- terms$end[cumsum(runs$lengths)]
  parties <- terms$party[cumsum(runs$lengths)]
  types <- terms$type[cumsum(runs$lengths)]
  
  data.frame(id=rep(id, length(starts)),
             firstName = rep(firstName, length(starts)),
             lastName = rep(lastName, length(starts)),
             gender = rep(gender, length(starts)),
             numTerms = runs$lengths,
             start = starts,
             end = ends,
             party = parties,
             type = types,
             stringsAsFactors = FALSE)
}

splinePoints <- function(theIndex, df){
  
  e <- df[theIndex,]
  
  # peak of the spline is the number of terms served
  peak <- e$numTerms

  # this output dataframe defines the shape of the spline. Lots of 
  # customization here.
  data.frame(x=quantile(c(e$start, e$end), probs=c(0,.15,.85,1)), 
             y=c(0, peak*.85, peak*.85, 0),
             splineIndex = e$id,
             party=e$party)
}


temp = lapply(ledge, extractElements)
ledge_df = bind_rows(temp)

ledge_df$start= ymd(ledge_df$start)
ledge_df$end <- ymd(ledge_df$end)
ledge_df$gender <- as.factor(ledge_df$gender)
ledge_df$party <- as.factor(ledge_df$party)
ledge_df$type <- as.factor(ledge_df$type)
ledge_df$dur <- ledge_df$end - ledge_df$start

ledge_df[ledge_df$party=="Republican",]$numTerms <- ledge_df[ledge_df$party=="Republican",]$numTerms * -1

```
```{r Senate, echo=FALSE}

body <- ledge_df %>% 
  filter(dur < 22000) %>%
  filter(type=="sen") %>% 
  filter(party %in% c("Democrat", "Republican"))%>% 
  arrange(start, end) 

# cheap and dirty way to move along a dataframe
temp <- lapply(X=seq_len(nrow(body)), splinePoints, body)
theSplines <- bind_rows(temp)


ggplot(theSplines, aes(x, y, group=splineIndex, colour=party)) + 
  geom_xspline(spline_shape = -0.5, size=0.2) +
  scale_color_manual("US Senate", values=c("blue", "red")) +
  theme_few() + theme(axis.text.y = element_blank(),
                      axis.title.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      panel.border = element_blank()) +
  xlab("Year")
```

## The Decline of American Politics

The Founding Fathers created the United States with the concept of citizen servants in mind. Men (originally, and land-owners at that) would represent constituents for a time and then return to their lives. Fresh-off the American Revolution, and having land to tend, this seemed like a good idea: who would want to spend time governing instead of working the land and being a member of the community where you built your life?

After being elected to two terms of the Presidency, George Washington decided to not run again, not wanting to create a democratic monarchy. Soon, the  government codified Presidential Term Limits in the US Constitution. Those term limits never trickled down to the Senate or the House; over time, legislators have spent more and more time legislating at the expense of being a member of their "home" communities.

```{r echo=FALSE}
ledge_df$numTerms = abs(ledge_df$numTerms)

body <- ledge_df %>% 
  filter(type=="sen") %>% 
  arrange(start, end) 

# cheap and dirty way to move along a dataframe
temp <- lapply(X=seq_len(nrow(body)), splinePoints, body)
theSplines <- bind_rows(temp)


ggplot(theSplines, aes(x, y, group=splineIndex)) + 
  geom_xspline(spline_shape = -0.5, size=0.2, colour="black") +
  theme_few() + theme(axis.text.y = element_blank(),
                      axis.title.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      panel.border = element_blank()) +
  xlab("Year")

```

This plot shows the service arcs of every Senator to have *finished* serving (that is, currently sitting Senators are not shown). The heights of the arcs are related to the number of terms served the the horizontal axis progresses by year to the right. The arcs tend to get higher over time, which indicates that Senators are serving longer. The horizontal striations - the *levels* - exactly correspond to the number of terms served: one term, two terms, etc.

```{r echo=FALSE}
bks = seq(trunc(min(year(body$end)) / 10) * 10, trunc(max(year(body$end)) / 10) * 10 + 10, 10)
body$decade = bks[cut(year(body$end), breaks=bks, labels=FALSE)]

decades <- body %>%
  group_by(decade) %>% 
  summarize(aveDur = mean(as.numeric(dur, units="days")/365),
            aveTerms = mean(numTerms)) 

ggplot(decades, aes(x=decade, y=aveTerms, colour=decade)) + 
  geom_point(colour="black") + geom_smooth(method="lm", se=F, linetype="dashed", size=0.5) +
  theme_few() + ylab("Average Number of Terms") +
  xlab("Decade in Which Service Expired") +
  theme(panel.border = element_blank()) 

```

The average of the number of terms for every Senator who finished servicing in a partcular decade clearly shows an up-and-to-the-right increase: the number of terms each Senator is serving is increasing.

## Party Time

```{r echo=FALSE, message=FALSE}
belows = c("Republican", "Jackson", "Unknown", "Adams", "American", "Populist")
ldf = ledge_df
ledge_df[ledge_df$party %in% belows,]$numTerms <- ledge_df[ledge_df$party %in% belows,]$numTerms * -1

body <- ledge_df %>% 
  filter(dur < 22000) %>%
  filter(type=="sen") %>% 
  arrange(start, end) 

# cheap and dirty way to move along a dataframe
temp <- lapply(X=seq_len(nrow(body)), splinePoints, body)
theSplines <- bind_rows(temp)

ggplot(theSplines, aes(x, y, group=splineIndex, colour=party)) + 
  geom_xspline(spline_shape = -0.5, size=0.2) +
  theme_few() + theme(axis.text.y = element_blank(),
                      axis.title.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      panel.border = element_blank()) +
  xlab("Year") + labs(colour="Party\nAffiliation")
```

American politics are dominated by political parties, and America has had lots of Senators from across a wide spectrum serve. Trying to break them out visually is difficult, and probably not all that important. While the "Liberal Republican", "Readjuster", and other parties have a place in American History; today, the landscape is dominated by two parties: "Democrat" and "Republican".

```{r echo=FALSE}
ledge_df = ldf

ledge_df[ledge_df$party=="Republican",]$numTerms <- ledge_df[ledge_df$party=="Republican",]$numTerms * -1

body <- ledge_df %>% 
  filter(dur < 22000) %>%
  filter(type=="sen") %>% 
  filter(party %in% c("Democrat", "Republican"))%>% 
  arrange(start, end) 

# cheap and dirty way to move along a dataframe
temp <- lapply(X=seq_len(nrow(body)), splinePoints, body)
theSplines <- bind_rows(temp)


ggplot(theSplines, aes(x, y, group=splineIndex, colour=party)) + 
  geom_xspline(spline_shape = -0.5, size=0.2) +
  scale_color_manual("US Senate", values=c("blue", "red")) +
  theme_few() + theme(axis.text.y = element_blank(),
                      axis.title.y = element_blank(),
                      axis.ticks.y = element_blank(),
                      panel.border = element_blank()) +
  xlab("Year")

```

Which brings us to the plot at the top of this post (replicated immediately above). It shows every Democrat and Republican Senator who has served since the inception of the Union (not including those currently serving).

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
